---
title: "R Notebook"
output: html_notebook
---


```{r message=FALSE}
library(tidyverse)
library(readr)
library(readxl)
library(plotly)
```



**Read in data**

```{r}
flight_num <- read_csv("Final Flights Data/Number of Flights(2015-2019).csv") %>% na.omit()
colnames(flight_num) <- tolower(gsub(" ", "_", colnames(flight_num)))
```
 
```{r}
flight_num %>% 
  pivot_longer(cols = "alaska_airlines" : "united_airlines", names_to = "flight", values_to = "number") -> flight_num2

flight_num2 %>% filter(year == 2019) %>% 
  ggplot() +
  geom_line(aes(month, number, color = flight), size = 1) +
  labs(title = "Number of Flights per Month in 2019")
```


```{r message=FALSE, warning=FALSE}
ggplot(flight_num2) +
  geom_line(aes(month, number, frame = year, color = flight)) +
  labs(title = "Number of Flights per Month per Year")-> p
ggplotly(p)
```
Southwest airlines have the most airlines over 2015 to 2019, and frontier_airlines is the second. The trend does not change much.

```{r}
flight_num2 %>% 
  group_by(flight, year) %>% 
  summarise(number = mean(number)) %>% 
  arrange(year, desc(number)) %>% pivot_wider(names_from = "year", values_from = number) 

```
**Delays**
```{r}
delays <- read_csv("Final Flights Data/Airline on-time perform(2015-2019).csv")  %>% 
  mutate()
colnames(delays) <- tolower(gsub(" ", "_", colnames(delays)))
```

delays due to what reason? 
```{r}
delays2 <- delays %>% pivot_longer(cols = "carrier_(avg._mins)" : "late_aircraft_arrival(avg._mins)", names_to = "reason", 
                        values_to = "minutes") 


delays2$minutes <- as.numeric(delays2$minutes)
na.omit(delays2)

reason <- delays2 %>% spread(key = reason, value = minutes, fill = 0) %>% 
  gather("carrier_(avg._mins)":"late_aircraft_arrival(avg._mins)", key = "reason", value = "minutes" ) %>% group_by(year, carriers, reason) %>% 
  summarise(minutes = mean(minutes)) %>% ungroup() -> reasons

reasons %>% group_by(year) %>% arrange(year, desc(minutes)) %>% mutate(rank = row_number())  %>% ungroup() %>% filter(rank <= 3) %>%
  ggplot() +
  geom_col(aes(year, minutes, fill = carriers), position = "dodge")
```

note: lack the information of Allegiant Airlines in 2015, 2016
delays %>% filter(carriers == "Allegiant Airlines") %>% View
for now: i just use na.omit


overall: 
```{r}
reasons %>% group_by(year) %>% arrange(year, desc(minutes)) %>% mutate(rank = row_number())  %>% ungroup() %>% filter(rank <= 3) %>% 
  ggplot() +
  geom_col(aes(year, minutes, fill = reason, color = carriers), position = "dodge", size = 1.5) +
  scale_fill_manual("legend", values = c("carrier_(avg._mins)" = "gray", "late_aircraft_arrival(avg._mins)" = "lightblue")) + 
  labs(title = "Delay Reasons") 
```
most delays over 5 years are due to carriers and late aircraft arrival
seems to be a common problem among all airlines

delay reason by year and airline
```{r}
reasons %>% group_by(year, carriers) %>% arrange(year, carriers, desc(minutes))
```




```{r}
# sample code
# movies %>% group_by(release_year, genre) %>%
#   count() %>% arrange(release_year, desc(n)) %>% 
#   spread(key = genre, value = n, fill = 0) %>% 
#   gather(2:16, key = "genre", value = "n") %>% 
#   arrange(release_year, genre) -> genre_year
# 
# ggplot(genre_year, aes(x = genre, y = n)) +
#   geom_point(aes(frame = release_year, color = genre, size = n)) + 
#   theme(axis.text.x = element_text(angle=65, vjust=0.5), legend.position = "none") -> ppn
# ggplotly(ppn)
```

